# Spatial prioritizations {#spatial-prioritizations}

## Introduction

Here we will develop prioritizations to identify priority areas for protected area establishment. Its worth noting that [prioritizr](https://prioritizr.net/), [Marxan](http://marxan.org/), and [Zonation](https://www.helsinki.fi/en/researchgroups/digital-geography-lab/software-developed-in-cbig#section-52992) are all decision _support_ tools. This means that these tools are all designed to _help_ you make decisions---they can't make decisions for you.

## Starting out simple

To start things off, let's keep things simple. Let's create a prioritization using the [minimum set problem formulation of the reserve selection problem](https://prioritizr.net/reference/add_min_set_objective.html). This problem will have 5% targets for each vegetation class and use the data in the `cost` column to specify acquisition costs. Although we strongly recommend using [Gurobi](https://www.gurobi.com/) to solve problems (with [`add_gurobi_solver`](https://prioritizr.net/reference/add_gurobi_solver.html)), we will use the [lpsymphony solver](https://prioritizr.net/reference/add_lsymphony_solver.html) in this workshop since it is much easier to install. The Gurobi solver is the fastest solver that prioritizr can use to generate solutions and it is much, much, much faster than the lpsymphony solver ([see here for Gurobi installation instructions](https://prioritizr.net/articles/gurobi_installation.html)).

```{r}
# print planning unit data
print(pu_data)

# make prioritization problem
p1 <- problem(pu_data, veg_data, cost_column = "cost") %>%
      add_min_set_objective() %>%
      add_relative_targets(0.05) %>% # 5% representation targets
      add_binary_decisions() %>%
      add_lpsymphony_solver(verbose = FALSE)

# print problem
print(p1)

# solve problem
s1 <- solve(p1)

# print solution,
# the solution_1 column contains the solution values with binary values
# indicating if a planning unit was selected as a (1) priority area or (0) not
print(s1)

# calculate number of planning units selected in the prioritization
sum(s1$solution_1)

# calculate total cost of the prioritization
sum(s1$solution_1 * s1$cost)

# plot solution
spplot(s1, "solution_1", col.regions = c("white", "darkgreen"), main = "s1")
```

Now let's examine the solution.

```{block2, type="rmdquestion"}
1. How many planing units were selected in the prioritization? What proportion of planning units were selected in the prioritization?
2. Is there a pattern in the spatial distribution of the priority areas?
3. Can you verify that all of the targets were met in the prioritization (hint: `feature_representation(p1, s1[, "solution_1"])`)?
4. What are limitations of this prioritization?

```

## Adding complexity

Our first prioritization suffers many limitations, so let's add additional constraints to the problem to make it more useful. First, let's lock in planing units that are already inside protected areas.

```{r}
# make prioritization problem
p2 <- problem(pu_data, veg_data, cost_column = "cost") %>%
      add_min_set_objective() %>%
      add_relative_targets(0.05) %>%
      add_locked_in_constraints("locked_in") %>%
      add_binary_decisions() %>%
      add_lpsymphony_solver(verbose = FALSE)

# print problem
print(p2)

# solve problem
s2 <- solve(p2)

# plot solution
spplot(s2, "solution_1", col.regions = c("white", "darkgreen"), main = "s2")
```

Let's pretend we talked to an expert on the vegetation communities in our study system and they recommended that a 20% target was needed. So, now let's set the targets to 20% of their total distribution in the study area.

```{r}
# make prioritization problem
p3 <- problem(pu_data, veg_data, cost_column = "cost") %>%
      add_min_set_objective() %>%
      add_relative_targets(0.2) %>%
      add_locked_in_constraints("locked_in") %>%
      add_binary_decisions() %>%
      add_lpsymphony_solver(verbose = FALSE)

# print problem
print(p3)

# solve problem
s3 <- solve(p3)

# plot solution
spplot(s3, "solution_1", col.regions = c("white", "darkgreen"), main = "s3")
```

Next, let's lock out highly degraded areas.

```{r}
# make prioritization problem
p4 <- problem(pu_data, veg_data, cost_column = "cost") %>%
      add_min_set_objective() %>%
      add_relative_targets(0.2) %>%
      add_locked_in_constraints("locked_in") %>%
      add_locked_out_constraints("locked_out") %>%
      add_binary_decisions() %>%
      add_lpsymphony_solver(verbose = FALSE)

# print problem
print(p4)

# solve problem
s4 <- solve(p4)

# plot solution
spplot(s4, "solution_1", col.regions = c("white", "darkgreen"), main = "s4")
```

Now, let's compare the solutions.

```{block2, type="rmdquestion"}
1. What is the cost of the planning units selected in `s2`, `s3`, and `s4`?
2. How many planning units are in `s2`, `s3`, and `s4`?
3. Do the solutions with more planning units have a greater cost? Why or why not?
4. Why does the first solution (`s1`) cost less than the second solution with protected areas locked into the solution (`s2`)?
5. Why does the third solution (`s3`) cost less than the fourth solution solution with highly degraded areas locked out (`s4`)?
6. Since planning units covered by existing protected areas have already been purchased, what is the cost for expanding the protected area system based on on the fourth prioritization (`s4`) (hint: total cost minus the cost of locked in planning units)?
7. What happens if you specify targets that exceed the total amount of vegetation in the study area? You can do this by rerunning the code to make `p4` with `add_absolute_targets(1000)` instead of `add_relative_targets(0.2)` and generating a new solution?

```

## Penalizing fragmentation

TODO.

## Budget limited prioritizations

In the real-world, the funding available for conservation is often very limited. As a consequence, decision makers often need prioritizations where the total cost of priority areas does not exceed a budget. In our fourth prioritization (`s4`), we found that we would need to spend an additional $`r round(sum(s4$cost * s4$solution_1) - sum(s4$cost * s4$locked_in))` million AUD to ensure that each vegetation community is adequately represented in the protected area system. But what if the funds available for protecting new land were limited to $1000? In this case, we need a "budget limited prioritization". Budget limited prioritizations aim to maximize some measure of conservation benefit subject to a budget (e.g. [number of species with at least one occurrence in the protected area system ](https://prioritizr.net/reference/add_max_cover_objective.html), or [phylogenetic diversity](https://prioritizr.net/reference/add_max_phylo_div_objective.html)). Let's create a budget-limited prioritization that aims to adequately represent as many targets as possible whilst remaining within a budget.

```{r}
# set the funds for additional land acquisition
funds <- 1000

# calculate the total budget for the prioritization
budget <- funds + sum(s4$cost * s4$locked_in)
print(budget)

# make prioritization problem
p6 <- problem(pu_data, veg_data, cost_column = "cost") %>%
      add_max_features_objective(budget) %>%
      add_relative_targets(0.2) %>%
      add_locked_in_constraints("locked_in") %>%
      add_locked_out_constraints("locked_out") %>%
      add_binary_decisions() %>%
      add_lpsymphony_solver(verbose = FALSE)

# print problem
print(p6)

# solve problem
s6 <- solve(p6)

# plot solution
spplot(s6, "solution_1", col.regions = c("white", "darkgreen"), main = "s5")

# find out which features have their targets met
r6 <- feature_representation(p6, s6[, "solution_1"])
print(r6$feature[r6$relative_held > 0.2])
```

We can also add weights to specify that it is more important to meet the targets for certain features and less important for other features. A common approach for weighting features is to assign a greater importance to features with smaller spatial distributions. The rationale behind this weighting method is that features with smaller spatial distributions are at greater risk of extinction. So, let's calculate some weights for our vegetation communities and see how weighting the features changes our prioritization.

```{r}
# calculate weights as the inverse number of grid cells that each vegetation
# class occupies, rescaled between 1 and 100
wts <- 1 / cellStats(veg_data, "sum")
wts <- rescale(wts, to = c(1, 10))

# print the name of the feature with smallest weight
names(veg_data)[which.min(wts)]

# print the name of the feature with greatest weight
names(veg_data)[which.max(wts)]

# plot histogram of weights
hist(wts, main = "feature weights")

# make prioritization problem with weights
p7 <- problem(pu_data, veg_data, cost_column = "cost") %>%
      add_max_features_objective(budget) %>%
      add_relative_targets(0.2) %>%
      add_feature_weights(wts) %>%
      add_locked_in_constraints("locked_in") %>%
      add_locked_out_constraints("locked_out") %>%
      add_binary_decisions() %>%
      add_lpsymphony_solver(verbose = FALSE)

# print problem
print(p7)

# solve problem
s7 <- solve(p7)

# plot solution
spplot(s7, "solution_1", col.regions = c("white", "darkgreen"), main = "s6")

# find out which features have their targets met when we add weights
r7 <- feature_representation(p7, s7[, "solution_1"])
print(r6$feature[r7$relative_held > 0.2])
```

```{block2, type="rmdquestion"}
1. What is the name of the feature with the smallest weight?
2. What is the cost of the sixth (`s6`) and seventh (`s7`) solutions?
4. Does there seem to be a big difference in which planning units were selected in the  sixth (`s6`) and seventh (`s7`) solutions?
3. Is there a difference between which features are adequately represented in the sixth (`s6`) and seventh (`s7`) solutions?

```

## Solution portfolios

In systematic conservation planning, only rarely do we have data on all of the stakeholder preferences and biodiversity features that we are interested in conserving. As a consequence, it is often useful to generate a portfolio of near optimal solutions to present to decision makers to guide the reserve selection process. Generally we would want many solutions in our portfolio (e.g. 1000) to ensure that our portfolio contains a range of highly different solutions, but here we will generate a portfolio containing just six near-optimal solutions so the code doesn't take too long to run. We will also increase the optimality gap to obtain solutions that are more suboptimal than earlier (the default gap value is 0.1).

```{r}
# make problem with a cuts portfolio
p8 <- problem(pu_data, veg_data, cost_column = "cost") %>%
      add_max_features_objective(budget) %>%
      add_relative_targets(0.2) %>%
      add_feature_weights(wts) %>%
      add_binary_decisions() %>%
      add_shuffle_portfolio(number_solutions = 6,
                            remove_duplicates = FALSE) %>%
      add_lpsymphony_solver(verbose = TRUE, gap = 10)

# print problem
print(p8)

# solve problem
# note that this will contain six solutions since we added a portfolio
s8 <- solve(p8)

# print solution
print(s8)
```

```{r, fig.width = 8, fig.height = 5}
# plot first solution
spplot(s8, "solution_1", col.regions = c("white", "darkgreen"),
       main = "s8 (solution 1)")

# plot all solutions
s8_plots <- lapply(paste0("solution_", seq_len(6)), function(x) {
  spplot(s8, x, main = x, col.regions = c("white", "darkgreen"))
})
do.call(grid.arrange, append(s8_plots, list(ncol = 3)))

# calculate the cost of the first solution
sum(s8$solution_1 * s8$cost)

# calculate the cost of the second solution
sum(s8$solution_2 * s8$cost)

# calculate the proportion of planning units with the same solution values
# in the first and second solutions
mean(s8$solution_1 == s8$solution_2)
```

```{block2, type="rmdquestion"}
1. What is the cost of each of the six solutions in portfolio? Are they costs very different?
2. Are the solutions in the portfolio very different?
3. What could we do to obtain a portfolio with more different solutions?

```
